---
title: "hw04-exercise"
output:
  github_document
---

# Tidy data and joins

## Initialize the data

* Load the gapminder, tidyverse and knitr libraries:

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
```

* We will take a quick look at the data to *sanity check* that the data and variables appear as we expect:

```{r}
(head(gapminder))
```

* Everything looks as expected, so let's start exploring the data

## Data Re-shaping Prompt
*Activity 2: Make a tibble with one row per year and columns for life expectancy for two or more countries. Use kable() to make the table look pretty. Scatterplot life expectancy for one country against that of another.*

### Method
* Reduce the data subset using select and group_by
* Spread the countries across the columns
* Build a table and a scatterplot

```{r}
q1_table <- gapminder %>% #Create a new variable for the tibble
  select(year, country, lifeExp) %>% #Reduce the data set
    group_by(year) %>% 
      spread(key = "country", value = "lifeExp") %>% #Spread countries across columns
        select("Canada", "Germany", "Brazil")# %>% #Select only two countries to display
        #summarize()

q1_table %>% 
  kable(col.names = c("Year", "Canada", "Germany", "Brazil"))

#Check that the data is a tibble

```
#Plot

```{r}
q1_table %>%
  ggplot(aes(x = year, y = value, color = variable)) +
  geom_point(aes(y = Canada, col = "Canada"), size=5) +
  geom_point(aes(y = Germany, col = "Germany"), size=5) +
  geom_point(aes(y = Brazil, col = "Brazil"), size=5) +
  #Add labels
  labs(title = "Life expectancy from 1952-2007 (Canada, Germany, Brazil)",
    x = "Year", y = "Life Expectancy",
    color = "Countries\n")

```


## Join Prompt (join, merge, look up)
*Activity 2: Make your own cheatsheet; iterate between your data prep and your joining to make your explorations comprehensive and interesting. Demonstrate all the types of joins.*

### Method
* Reduce the data subset using select and group_by
* Find the minimum and maximum values

```{r}

```


*Activity 3: Explore the merge() and match() functions. Contrast and compare.*

### Method
* Reduce the data subset using se


